Information within words in English, French and German speech

### 2019-07-01 surprisal/letter_surprisal.Rmd
phoneme writeup counts and edits

adult_utterances <- prov_utterances %>%
  filter(speaker_role != "Target_Child") %>% 
  mutate(utterance_id = 1:n()) %>% 
  unnest_tokens(word, gloss, token = stringr::str_split, pattern = "[ +]+") %>%
  group_by(utterance_id) %>%
  mutate(word_id = 1:n()) %>%
  ungroup() %>%
  unnest_tokens(letter, word, token = stringr::str_split, pattern = "", drop = F) %>%
  group_by(utterance_id, word_id) %>%
  mutate(letter_order = 1:n()) %>%
  mutate(word_length = max(letter_order)) %>%
  ungroup()
```

>>>>

french_utterances <- prov_utterances %>%
  filter(speaker_role != "Target_Child") %>% 
  mutate(utterance_id = 1:n()) %>% 
  unnest_tokens(word, gloss, token = stringr::str_split, pattern = "[ +]+") %>%
  group_by(utterance_id) %>%
  mutate(word_id = 1:n()) %>%
  ungroup() %>%
  unnest_tokens(letter, word, token = stringr::str_split, pattern = "", drop = F) %>%
  group_by(utterance_id, word_id) %>%
  mutate(letter_order = 1:n()) %>%
  mutate(word_length = max(letter_order)) %>%
  ungroup()

phone_counts <- adult_phonemes %>% 
  group_by(utterance_id, word_id) %>% 
  summarise(length = max(word_length)) %>% 
  ungroup() %>% 
  count(length)
```

### 2019-07-01 surprisal/phoneme_writeup.Rmd
phoneme writeup counts and edits

---
title: "Phoneme surprisal"
author: Josef Klafka and Dan Yurovsky 
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: false
    number_sections: false
    theme: lumen
    toc_float: false
    code_folding: hide 
editor_options: 
  chunk_output_type: console
---

@@

Where is the information in English words? A salient feature of African American Enlgish Vernacular is the reduction of final consonants and consonant clusters. Why must this happen? One possibility is that the information important for identifying a word is not uniformly distributed across it's characters or sounds but instead concentrated at the beginnings or middles of words. In these cases, reducing final sounds might reduce the articulatory complexity of words without much cost for the listener in processing these words and inferring their intended meaning.

In a series of initial analyses we tackle three ways:
1. Computing the surprisal of characters in English words either from baserates of character production (unigram) or conditioning on the prior charaters in each word (bigram and trigram). This analyses roughly captures the difficulty that a listener might have in predicting each character.
2. Repeating this analysis at the level of phonemes rather than characters, arguably a better proxy for the speaker's and listener's representations
3. Computing hold-out entropy for individual phonemes in English words. This is roughly equivalent to asking "if you knew every other sound in a word, how much trouble would you have predicting what sounds appears in position X"
4. Comparing English to French and German which have different expected information profiles (more below).

SOME KIND OF PREVIEW OF RESULTS HERE?

```{r read-data}
prov_letters <- read_feather(here("Data/providence_letters.feather")) %>%
  mutate(rep = "letter", language = "English") %>%
  rename(order = letter_order)

prov_phones <- read_feather(here("Data/providence_phonemes.feather")) %>%
  mutate(rep = "phone", language = "English") %>%
  rename(order = phone_order)

adult_phonemes <- read_feather(here("Data/adult_phonemes.feather"))

french_letters <- read_feather(here("Data/palasis_letters.feather")) %>%
  mutate(rep = "letter", language = "French") %>%
  rename(order = letter_order)

german_letters <- read_feather(here("Data/wagner_letters.feather")) %>%
  mutate(rep = "letter", language = "German") %>%
  rename(order = letter_order)

all_data <- bind_rows(prov_letters, prov_phones, french_letters, german_letters)
```

@@

We use for our English corpus the Providence corpus from CHILDES. In the plots below, each plot represents the information in each letter in words of different lengths by number of letters--we show the length above each plot.

We see a distinctive shape for information at the level of individual letters in English. The first letter has relatively high information, the second letter has less information and the third letter has high infomration, and the final letter (if the word has more than 3 letters) has less information.

```{r plot-helper}
plot_curves <- function(select_language = "English", select_rep = "letter", 
                        select_gram = "Unigram") {
  all_data %>%
    filter(language == select_language, rep == select_rep,
           gram == select_gram) %>%
    filter(word_length %in% 2:6) %>%
    ggplot(aes(x = order, y = mean_s)) +
    facet_grid(. ~ word_length, 
               labeller = as_labeller(function(x) paste0("Length ", x))) +
    geom_point() +
    geom_line() +
    xlab(paste0(str_to_sentence(select_rep), " Position")) +
    ylab("Surprisal") + 
    ggtitle(paste0("Information in ", select_language, " ", 
                   select_rep, "s by ",select_rep, " position - ", select_gram))
}
```

@@

## Surprisal

Briefly, what we're doing is splitting each word in the Providence corpus from CHILDES up into its constituent phones, i.e. the sounds that make up the word. We then calculate how common each phone is across the entire corpus. From this frequency measure, we compute the surprisal of each phone, which is inversely proportional to the frequency of the phone. The less common a phone is, the more difficult it is to process the phone.

We split up the words in the corpus by their length in number of phones--for example, the word "food" has three phones: one for the "f", one for the long "oo", and one for the "d". We would take all words like this that have three phones. For each phone position  within the words of that length, we calculate the average surprisal/information in that position. For example, looking at all of the words with three phones, we take all of the first phones and calculate how much information is in that position on average. In other words, how difficult is it to process the first sound in a word, compared to the other sounds in a word?

Unlike with letters, there isn't as much of a consistent shape to the information distribution for phones considered without context. For the most part, the first phone has relatively little information while the second phone has a lot of information.

@@

## Hold-out entropy

For this next analysis, we took all words with 2, 3, 4, 5 or 6 phones. For each word length, we held out the phones in one phone position. For example, in all words with 4 phones, we held out the first phone. We then computed the entropy, or amount of variation, in the remaining words. The higher the entropy when holding out phones in a given position, the more information is contained in that phone on average.

@@

We are interested in French because when many French words are pronounced, the final letters are not pronounced in any way (e.g. "ils sauraient"="they would know", the final "ent" in "sauraient" is not pronounced at all). We are interested in German because German has a much richer case system than English or French, and case markings in German are suffixes (at the ends of words). So in German, the ends of words may be much more important than in French or English.

# French letters

@@

# German letters

>>>>

---
title: "Information within words in English, French and German speech"
author: Josef Klafka and Dan Yurovsky 
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: false
    number_sections: true
    theme: lumen
    toc_float: false
    code_folding: hide 
editor_options: 
  chunk_output_type: console
---

@@

```{r read-data}
all_data <- read_feather(here("Data/all_phonemes.feather"))
all_counts <- read_feather(here("Data/all_counts.feather"))
```

How is information spread among the sounds in English words? A salient feature of African American English Vernacular is the reduction of final consonants and consonant clusters. Why hasn't this created communication problems, namely identifying and disambiguating words? One possibility is that the information important for identifying a word is not uniformly distributed across all of its characters or sounds, but instead concentrated at the beginnings or middles of words. In these cases, reducing final sounds might reduce the articulatory complexity of words without much cost for the listener in processing these words and inferring their intended meaning. By the time the listener hears only the first sounds of a word, they may already have identified which word they are hearing.

Through a series of exploratory analyses we tackle this problem in three ways:
1. Computing the surprisal of characters in English words from baserates of character production (unigram) and then conditioning on the prior charaters in each word (bigram and trigram). This analyses roughly captures the difficulty that a listener might have in predicting each character.
2. Repeating this analysis at the level of phonemes rather than characters, arguably a better proxy for the speaker's and listener's representations of spoken English. 
3. Computing hold-out entropy for individual phonemes in English words. This is roughly equivalent to asking "if you knew every other sound in a word, how much trouble would you have predicting what sounds appears in position $X$". 
4. Comparing our English by-letter analysis to results for French and German which have different expected information profiles. We expect that French words will have less information at the end of words compared to English, while German words will have more compared to English.

Across all of our analyses, we find that the letters and sounds at the end of words in English have less information than the first letters and sounds in a word. Letters at the end of words in French and German also have little information compared to letters at the beginning of words in those languages, like in English.

@@

We use the Providence corpus from CHILDES as data for English. In the plots below, each plot represents the information in each letter, divided by words of different lengths. Word lengths are given by number of letters--we show the length above each plot.

We see a distinctive shape for information at the level of individual letters in English. The first letter has relatively high information, the second letter has less information and the third letter has high infomration, and the final letter has less information (if the word has more than $3$ letters).

```{r plot-helper}
plot_curves <- function(select_language = "English", select_rep = "letter", 
                        select_gram = "Unigram") {
  all_data %>%
    filter(language == select_language, rep == select_rep,
           gram == select_gram) %>%
    filter(word_length %in% 2:6) %>%
    ggplot(aes(x = order, y = mean_s)) +
    facet_grid(. ~ word_length, 
               labeller = as_labeller(function(x) paste0("Length ", x, " \n n = ", 
                                                         all_counts %>% 
                                                           filter(language == select_language,
                                                                  rep == select_rep, 
                                                                  length == x) %>%
                                                           pull(length)))) +
    geom_point() +
    geom_line() +
    xlab(paste0(str_to_sentence(select_rep), " Position")) +
    ylab("Average information") + 
    ggtitle(paste0("Information in ", select_language, " ", 
                   select_rep, "s by ",select_rep, " position - ", select_gram))
}
```

@@

Briefly, what we're doing is splitting each word in the Providence corpus from CHILDES up into its constituent phones--the sounds that make up the word. We split up the words in the corpus by their length in number of phones--for example, the word "food" has three phones: one for the "f", one for the long "oo", and one for the "d". We would take all words like this that have three phones. For each phone position  within the words of that length, we calculate the average information in that position.

Unlike with letters, there isn't as much of a consistent shape to the information distribution for phones in the unigram condition. If anything, the first phone has relatively little information while the second phone has a lot of information.

@@

# Hold-out entropy

For this next analysis, we took all words with $2$, $3$, $4$, $5$ or $6$ phones. For each word length, we held out the phones in one phone position. For example, in all words with 4 phones, we held out the first phone. We then computed the entropy, or amount of variation, in the remaining words. The higher the entropy when holding out phones in a given position, the more information is contained in that phone on average.

@@

# French and German

For our final analysis, we're going to look at how the information within words in French and German compare to English. We use the Palasis French corpus and Wagner German corpus from CHILDES for our data. We chose French because for many French words the final letters are not pronounced regardless of how the letters in the word appear. For example, "ils sauraient" meaning "they would know" is pronounced without the final "ent" in "sauraient". We are interested in German because German has a much richer case system than English or French, where case markings in German are suffixes at the ends of words. We think that in German, the ends of words may be more important than in French or English.

## French letters

@@

## German letters
